---
title: "Assignment 9"
author: "Jaclyn Narleski"
date: "2025-03-18"
output: pdf_document
---

# Setup

```{r}

# Load necessary packages
library(tidyverse)
library(here)
library(naniar)
library(DHARMa)
library(gtsummary)
library(broom)

# Read csvs
lizards <- read.csv("jrn_lizard.csv")
lizard_npp <- read.csv("jrn_lizard_npp.csv")

```

# Logistic Regression

Do snout-to-vent length, sex, and vegetation zone at time of capture significantly predict if a lizard tail is recorded as whole?

## Data Tidying

```{r}

# Filter dataset
lizards <- lizards %>%
  filter(spp == "UTST")

```

## Examine Data

```{r}

# Create exploratory figures for each variable
(hist1 <- ggplot(data = lizards, aes(x = SV_length)) +
    geom_histogram() +
    theme_bw())

(hist2 <- ggplot(data = lizards, aes(x = sex)) +
    geom_histogram(stat = "count", na.rm = TRUE) +
    theme_bw())

(hist3 <- ggplot(data = lizards, aes(x = zone)) +
    geom_histogram(stat = "count", na.rm = TRUE) +
    theme_bw())

(hist4 <- ggplot(data = lizards, aes(x = tail)) +
    geom_histogram(stat = "count", na.rm = TRUE) +
    theme_bw())

# Create exploratory figures between variables
(hist5 <- ggplot(data = lizards, aes(x = SV_length,
                                            fill = tail)) +
    geom_histogram(na.rm = TRUE) +
    theme_bw())

(hist6 <- ggplot(data = lizards, aes(x = sex,
                                            fill = tail)) +
    geom_histogram(stat = "count", na.rm = TRUE) +
    theme_bw())

(hist7 <- ggplot(data = lizards, aes(x = zone,
                                            fill = tail)) +
    geom_histogram(stat = "count", na.rm = TRUE) +
    theme_bw())

```

## Fit Regression Model

```{r}

# Turn character vectors into factors
lizards$sex <- factor(lizards$sex, 
                      levels = c("F", "J", "M", "NA"))

lizards$zone <- factor(lizards$zone,
                       levels = c("C", "G", "M", "T"))

lizards$tail <- factor(lizards$tail,
                       levels = c("B", "W", "NA"))

# Fit regression model
tail_mod <- glm(tail ~ SV_length + zone + sex, 
                data = lizards,
                family = "binomial")

# Summary of regression model
summary(tail_mod)

# Simulate residuals in plot form 
simulateResiduals(tail_mod) %>% plot()

```
## Create Prediction Model

```{r}

# Create vector of snout-to-vent lengths
svlength_vector <- rep(seq(from = 0, to = 50), 4)

# Vector of sex categories
sex_vector <- c(rep("F", 51),
                 rep("M", 51),
                 rep("J", 51),
                 rep(NA, 51))

# Vector of zone categories
zone_vector <- c(rep("C", 51),
                 rep("G", 51),
                 rep("M", 51),
                 rep("T", 51))

# Generate data frame
data_pred <- data.frame(svlength_vector, sex_vector, zone_vector)
# Name columns
colnames(data_pred) <- c("SV_length", "sex", "zone")

# Generate prediction model
prediction <- predict(tail_mod,
                      newdata = data_pred,
                      type = "response",
                      se.fit = TRUE)

# Generate data frame of predictions
data_fig <- data.frame(data_pred,
                       prediction$fit,
                       prediction$se.fit)
# Name columns
colnames(data_fig) <- c("SV_length", 
                        "Sex", "Zone", "probability", "se")

# Generate plot for with zone coloring
zoneplot <- ggplot(data_fig, aes(x = SV_length,
                     y = probability)) +
  geom_line(aes(color = Zone)) +
  geom_ribbon(aes(ymin = probability - se,
                  ymax = probability + se,
                  fill = Zone), alpha = 0.3) +
  labs(x = "Snout-to-Vent Length",
       y = "Probability of Whole Tail",
       color = "Zone",
       fill = "Zone") +
  scale_color_manual(values = c("chartreuse3", "darkolivegreen3", "yellow3", 
                                "palegreen3"),
                     labels = c("Black Grama Grassland", 
                                "Creosotebush Shrubland",
                                "Mesquite Duneland", "Tarbush Shrubland")) +
  scale_fill_manual(values = c("greenyellow", "darkolivegreen", "yellow4", 
                                "palegreen3"),
                    labels = c("Black Grama Grassland", 
                               "Creosotebush Shrubland",
                               "Mesquite Duneland", "Tarbush Shrubland")) +
  theme_bw()

# View plot
zoneplot

# Generate plot for sex coloring
sexplot <- ggplot(data_fig, aes(x = SV_length,
                                y = probability)) + 
  geom_line(aes(color = Sex)) +
  geom_ribbon(aes(ymin = probability - se,
                  ymax = probability + se,
                  fill = Sex), alpha = 0.3) +
  labs(x = "Snout-to-Vent Length",
       y = "Probability of Whole Tail",
       color = "Sex",
       fill = "Sex") +
    scale_color_manual(values = c("violetred3", "purple3", "gold3", 
                                "snow3"),
                     labels = c("Female", "Male","J", "No Data")) +
    scale_fill_manual(values = c("violetred3", "purple3", "gold3", 
                                "snow3"),
                     labels = c("Female", "Male","J", "No Data")) +
  theme_bw()

# View plot 
sexplot
```
## Intepret Models

The results of a logistic regression suggest that the odds of a lizard having a whole tail are greater when the snout-to-vent length is lower (beta = -0.04523, p < 0.01) and the lizard is less likely to be located in the black grama grassland (beta = -0.56537, p = 0.001) or the tarbush shrubland (beta = -0.40435, p < 0.05) zones.

# Poisson Regression

Do season and plant species percent cover significantly predict lizard counts?

## Examine Data

```{r}

# Generate exploratory plots for each visual
(hist8 <- ggplot(data = lizard_npp, aes(x = season)) +
    geom_histogram(stat = "count", na.rm = TRUE) +
    theme_bw())

(hist9 <- ggplot(data = lizard_npp, aes(x = BOER)) +
    geom_histogram(na.rm = TRUE) +
    theme_bw())

(hist10 <- ggplot(data = lizard_npp, aes(x = LATR)) +
    geom_histogram(na.rm = TRUE) +
    theme_bw())

(hist11 <- ggplot(data = lizard_npp, aes(x = PRGL)) +
    geom_histogram(na.rm = TRUE) +
    theme_bw())

(hist12 <- ggplot(data = lizard_npp, aes(x = SCBR)) +
    geom_histogram(na.rm = TRUE) +
    theme_bw())

(hist13 <- ggplot(data = lizard_npp, aes = lizard_count)) +
  geom_histogram(na.rm = TRUE)

```

## Fit Regression Model

```{r}

# Make season characters into factors
lizard_npp$season <- factor(lizard_npp$season, 
                      levels = c("F", "W", "S"))

# Fit regression model
lizard_nppmod <- glm(lizard_count ~ season + BOER + LATR + PRGL + SCBR,
                data = lizard_npp,
                family = "poisson")

# View summary
summary(lizard_nppmod)
# View plots
plot(lizard_nppmod)

# Remove more outlier rows
lizard_npp2 <- lizard_npp[-c(32, 47, 50, 1),]

# Fit other regression model(s) and view summary/plots
lizard_nppmod2 <- glm(lizard_count ~ season + BOER + LATR + PRGL + SCBR,
                data = lizard_npp2,
                family = "poisson")


summary(lizard_nppmod2)
plot(lizard_nppmod2)

lizard_nppmod3 <- glm(lizard_count ~ season + PRGL + SCBR,
                data = lizard_npp2,
                family = "poisson")

summary(lizard_nppmod3)
plot(lizard_nppmod3)

lizard_nppmod4 <- glm(lizard_count ~ season + BOER + PRGL + SCBR,
                data = lizard_npp2,
                family = "poisson")

summary(lizard_nppmod4)
plot(lizard_nppmod4)

# This fourth model has the best significance and AIC

# View regression p-values and confidence intervals 
tbl_regression(lizard_nppmod4)

```

## Interpret Models

The results of a Poisson regression show that lizard counts are lower in the summer (beta = -0.82462, p < 0.01), lower in BOER (beta = -0.05221, p < 0.05), and higher in PRGL (beta = 0.13224, p < 0.01) an in SCBR (beta = 0.43805, p < 0.01). 
